---
title: "lab03-functions-vectors"
author: "Joseph Chamdani, Winson Teh, Kenneth Wu, Hanjun Ma"
date: "2025-10-14"
output: html_document
---

## **1 Loops**

---

### 1.1

Write a for-loop that finds all such numbers among the first 300 integers.

```{r}
for (n in 1:300) {
  if ((1/n) * n != 1) {
    cat(n, " ")
  }
}
```

---

### 1.2

Make a function that expects two arguments, n1 and n2, and prints all such numbers between n1 and n2.

```{r}
errors <- function(n1, n2) {
  for (n in n1:n2) {
    if ((1/n) * n != 1) {
      cat(n, " ")
    }
  }
  cat("\n")  
}

errors(500, 600)
```

---

### 1.3

Add default value for the argument n1: if not specified, it should be 1. Show that your function works.

```{r}
errors <- function(n1 = 1, n2) {
  for (n in n1:n2) {
    if ((1/n) * n != 1) {
      cat(n, " ")
    }
  }
  cat("\n")  
}
```

```{r}
errors(n2 = 200)

errors(100, 200)
```

---

## **2 Vectors**

---

### 2.1 
Create string vector of gods: "Baldur", "Yggdrasil", "Odin", "Loki". Give it an appropriate name.

```{r}
gods <- c("Baldur", "Yggdrasil", "Odin", "Loki")
gods
```
---

### 2.2
Convert all god names to upper case.

```{r}
gods_upper <- toupper(gods)
gods_upper

```
---

## 2.3
Make the vector of god names into a single string: “Baldur and Yggdrasil and Odin and Loki”.

```{r}
gods_string <- paste(gods, collapse = " and ")
gods_string
```

---

### 2.4
Use indexing to separate the first three gods (Baldur–Odin) from the gods list and store in a different vector.

```{r}
first_three_gods <- gods[1:3]
first_three_gods
```

---

### 2.5
Combine these three gods into a single string “Baldur, Yggdrasil, Odin”. (use paste as above)

```{r}
first_three_gods_string <- paste(first_three_gods, collapse = ", ")
first_three_gods_string
```

---

## 2.6 
Now use the previous string and the last god’s name to make a vector “Baldur, Yggdrasil, Odin and Loki”

```{r}
all_gods_string <- paste(first_three_gods_string, "and", gods[4])
all_gods_string
```

---

### 2.7
Now achieve the same using function str_flatten_comma from the stringr library. Look up the help for the function yourself!

```{r}
library(stringr)

all_gods_str <- str_flatten_comma(gods, last = " and ")
all_gods_str
```

## **3 Logical indexing**

```{r}
country <- c("Mozambique", "Kuwait", "Colombia","Tunisia", "Dominican Republic", "Gambia")
continent <- c("Africa", "Asia", "Americas","Africa", "Americas", "Africa")
population <- c(34174, 4310, 52085,12458, 11333, 2773) # in thousands
capital <- c("Maputo", "Kuwait City", "Bogotá","Tunis", "Santo Domingo", "Banjul")
```


---

### 3.1 
How many countries are there with more than 10M inhabitants (i.e. the population in thousands larger than 10,000)?

```{r}
more_than_10000_countries <- population > 10000
num_more_than_10000_countries <- sum(more_than_10000_countries)
num_more_than_10000_countries
```

---

### 3.2
Print population of all such countries

```{r}
population[more_than_10000_countries]
```

---

### 3.3 
Compute a vector of length 2 that contains a) number of “small countries” (no more than 10M inhabitants) and b) number of “large countries” (at least 10M inhabitants).

```{r}
small_countries <- sum(population <= 10000)
large_countries <- sum(population > 10000)

country_populations <- c(small_countries,large_countries)
names(country_populations) <- c("small", "large")
country_populations
```

---

### 3.4
Names of all large countries

```{r}
country[more_than_10000_countries]
```

---

### 3.5 
Capitals of all countries in Africa

```{r}
africa_country <- continent == "Africa"
capital[africa_country]
```

---

### 3.6
Capitals of large countries in Americas

```{r}
americas_country <- continent == "Americas" & more_than_10000_countries
capital[americas_country]
```

---

### 3.7 
Which countries have less than 10M inhabitants in Asia?

```{r}
less_than_10000_countries <- population < 10000
asia_small <- continent == "Asia" & less_than_10000_countries
country[asia_small]
```


