---
title: "PS2"
author: "Joseph Davis Chamdani"
date: "2025-10-14"
output: html_document
---

## **1 Vectors and vectorized operations**

---

### **1.1 Create vectors**

#### **1.1.1 Six movies you like**

```{r}
movies <- c("Iron Man", "Superman", "Ant Man",
            "Interstellar", "The Dark Knight", "Whiplash")
movies
```
---

#### **1.1.2 First three movies (indexing)**
```{r}
first_three <- movies[1:3]
first_three
```

---

#### **1.1.3 Add "is a great movie!" to each element (string ops)**
```{r}
movies_great <- paste(movies, " is a great movie!")
movies_great
```

---

#### **1.1.4 All movies except the 4th (indexing)**
```{r}
movies_no4 <- movies[-4]
movies_no4
```
---

#### **1.1.5 Numbers 40 through 49 (inclusive)**
```{r}
numbers <- 40:49
numbers
```
---

#### **1.1.6 How many numbers? (length)**
```{r}
len_numbers <- length(numbers)
len_numbers
```
---

#### **1.1.7 Mean of numbers**
```{r}
mean_numbers <- mean(numbers)
mean_numbers
```

---

#### **1.1.8 Another vector: 70 through 79 (inclusive)**
```{r}
more_numbers <- 70:79
more_numbers
```

---

#### **1.1.9 Combine both number vectors**
```{r}
all_numbers <- c(numbers, more_numbers)
all_numbers
```

---

### **1.2 Logical Indexing**

#### **1.2.1 Create logical vector: numbers larger than 70**

```{r}
iLogical <- all_numbers > 70
iLogical
```
The TRUE/FALSE means that if it's true it's above 70 if it's false it's below 70.

---

#### **1.2.2 Compute sum of iLogical**

```{r}
sum_iLogical <- sum(iLogical)
sum_iLogical
```
---

#### **1.2.3 Use iLogical variable to extract only those numbers that are larger than 70**

```{r}
larger_than_70 <- all_numbers[iLogical]
larger_than_70
```

---

#### **1.2.4 Check if they look correct**
```{r}
larger_than_70
```

---

#### **1.2.5 Compute the length of this subset**
```{r}
length(larger_than_70)
```

---

#### **1.2.6 Use logical NOT (!) to extract numbers that do NOT exceed 70**

```{r}
not_exceed_70 <- all_numbers[!iLogical]
not_exceed_70
```

---

#### **1.2.7 Print squares of numbers that do not exceed 70**

```{r}
not_exceed_70^2
```
---

#### **1.2.8 Compute and print squares of all numbers that do not exceed 70**

```{r}
is_even <- all_numbers %% 2 == 0
is_even
```
---

#### **1.2.9 Use logical indexing to print only even numbers**

```{r}
even_numbers <- all_numbers[is_even]
even_numbers
```

---

#### **1.2.10 Between 1 and 200: 1/n * n ≠ 1 (since 1/n * n = 1 for all n except 0)**

```{r}
n <- 1:200
logic_n <- (1/n * n) != 1
logic_n
```

--- 

#### **1.2.11 Count how many such numbers you have between 1 and 1000**

```{r}
n2 <- 1:1000
logic_n2 <- (1/n2 * n2) != 1
sum(logic_n2)
```
___

#### **1.2.12 Extra Credit (Python) **

```{python}
numbers = range(1, 201)
wrong_nums = [n for n in numbers if (1/n) * n != 1]
print(wrong_nums)
print("Count:", len(wrong_nums))
```

---

### **1.3 Modifying vectors**

#### **1.3.1 Create a vector of numbers −5, −4, . . . , 4, 5.**

```{r}
nums <- -5:5
nums
```

---

#### **1.3.2 Replace all negative numbers with “0”**

```{r}
nums <- -5:5
nums[nums < 0] <- 0 
nums
```

---

#### **1.3.3 Now implement ReLU function:**

```{r}
relu <- function(x) {
  x[x < 0] <- 0
  return(x)
}
```

---

#### **1.3.4 Test ReLU with example vectors**

```{r}
relu(c(-2, -1, 0, 1, 2))
relu(c(-10, 5, -3, 7, 0))
```

---

#### **1.3.5 Implement absolute value function (custom)**

```{r}

```

---










